---
title: "Investigating the 2021 Texas Power Crisis"
subtitle: "EDS-223 Assignment 3"
author: "Isabella Segarra"
date: 10-19-2025
format: 
  pdf: default 
  html: default
theme: pandoc
execute: 
  warning: false
  message: false
---

# TITLE 

[ADD README SCREENSHOT]

### Objective
In this project, I investigated....

## Environment Set-up 
```{r}
#| output: false
#| message: false

# Load relevant libraries
library(sf) # For vector data
library(stars) # For raster data 
library(tmap) # For static and interactive maps
library(here) # For importing data 
library(tidyverse) # For data cleaning
library(dplyr) # For filtering data 
library(paletteer) # For pretty colors 
library(testthat) # For efficient workflows 

```

## Data Import
The data I am using in this project includes data from NASA's Worldview for night light data, data for roads and houses from OpenStreetMap (OSM), and socieconomic data from the U.S. Census Bureau's American Community Survey. 

### Night lights data import
This raster data is from NASA's Worldview tool, distributed through NASA's Level-1 and Atmospheric Archive & Distribution System Distributed Active Archive Center (LAADS DAAC). It is a `.tif` file that contains 10x10 degree tiles. I will be importing four data files from the `VNP46A1` folder for night light data from the days surrounding the storm. 
```{r}

# Import tile h08v05 from 2021-02-07
nl_h08v05_07 <- read_stars(here("data", "VNP46A1","VNP46A1.A2021038.h08v05.001.2021039064328.tif"))

# Import tile h08v06 from 2021-02-07
nl_h08v06_07 <- read_stars(here("data", "VNP46A1","VNP46A1.A2021038.h08v06.001.2021039064329.tif"))

# Import tile h08v05 from 2021-02-16
nl_h08v05_16 <- read_stars(here("data", "VNP46A1","VNP46A1.A2021047.h08v05.001.2021048091106.tif"))

# Import tile h08v06 from 2021-02-16
nl_h08v06_16 <- read_stars(here("data", "VNP46A1","VNP46A1.A2021047.h08v05.001.2021048091106.tif"))

```

### Roads data import
This vector data is from OSM, processed with the help of a third party site (CITE) to produce a shapefile of all Texas highways. This data contains 58,6056 observations with 11 variables. 
```{r}
# Import roads data
roads <- read_sf(here("data", "gis_osm_roads_free_1.gpkg"))

```

### House data import
This vector data is also from OSM, processed with the helpf of a third party(CITE) to produce a shapefile of houses in the Houseton metropolitan area. This data contains 49,8795 observations with 6 variables. 
```{r}
# Import house data 
houses <- read_sf(here("data", "gis_osm_buildings_a_free_1.gpkg"))

```

## Socieconomic data import
This data is from  U.S. Census Bureau’s American Community Survey (CITE) 2019 census. This Geodatabase (gdb) data contains 5,365 observations with 719 variables.  
```{r}
# Import socioeconomic data REDO!!!
socioeco <- read_sf(here("data", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"))


```

```{r}
print(st_crs(houses))
print(st_crs(roads))
print(st_crs(socioeco))
```


## Data wranlging

### CRS Matching
In order to work with all datasets, I need to confirm that the Coordinate Refrence Systems (CRS) match. 

```{r}
# ......Check CRS of individual datasets ......
print(st_crs(houses)$epsg) 
print(st_crs(roads)$epsg)
print(st_crs(socioeco)$epsg)

# Night light data 
print(st_crs(nl_h08v05_07)$epsg)
print(st_crs(nl_h08v06_07)$epsg)
print(st_crs(nl_h08v05_16)$epsg)
print(st_crs(nl_h08v06_07)$epsg)

```

```{r}

if(st_crs(houses) == st_crs(roads)) {
  print("coordinate reference systems match")
} else{
  print("coordinate reference systems do NOT match")
} 

if(st_crs(roads) == st_crs(socioeco)) {
  print("coordinate reference systems match")
} else{
  print("coordinate reference systems do NOT match")
} 

if(st_crs(socieco) == st_crs(socioeco)) {
  print("coordinate reference systems match")
} else{
  print("coordinate reference systems do NOT match")
} 
```

## Phase 1: Create blackout mask 
In this phase, I will identifyt places that experienced a blackout by creating a mask. 
```{r}

```

