---
title: "Investigating the 2021 Texas Power Crisis"
subtitle: "EDS-223 Assignment 3"
author: "Isabella Segarra"
date: 10-19-2025
format: 
  pdf: default 
  html: default
theme: pandoc
execute: 
  warning: false
  message: false
---

# TITLE 

[ADD README SCREENSHOT]
![](filepath)

### Objective
In this project, I investigated the February 2021 Texas power crisis, which was triggered by three severe winter storms that caused widespread blackouts across the region. To illustrate the impact of the storms, I mapped Houston’s night lights before and after the crisis and identified areas that experienced a blackout. To explore the differential effects of the power outages from a socioeconomic perspective, I compared the median household incomes of census tracts that experienced a blackout with those that did not. 

## Environment Set-up 
```{r}
#| output: false
#| message: false

# Load relevant libraries
library(sf) # For vector data
library(stars) # For raster data
library(terra) # raster handling
library(tmap) # For static and interactive maps
library(here) # For importing data 
library(tidyverse) # For data cleaning
library(dplyr) # For filtering data 
library(paletteer) # For pretty colors 
library(testthat) # For efficient workflows 

```

## Data Import
The data I am using in this project includes data from NASA's Worldview for night light data, data for roads and houses from OpenStreetMap (OSM), and socieconomic data from the U.S. Census Bureau's American Community Survey. 

### Night lights data import
This raster data is from NASA's Worldview tool, distributed through NASA's Level-1 and Atmospheric Archive & Distribution System Distributed Active Archive Center (LAADS DAAC). It is a `.tif` file that contains 10x10 degree tiles. I will be importing four data files from the `VNP46A1` folder for night light data from the days surrounding the storm as stars. 
```{r}

# Import tile h08v05 from 2021-02-07
nl_h08v05_07 <- read_stars(here("data", "VNP46A1","VNP46A1.A2021038.h08v05.001.2021039064328.tif"))

# Import tile h08v06 from 2021-02-07
nl_h08v06_07 <- read_stars(here("data", "VNP46A1","VNP46A1.A2021038.h08v06.001.2021039064329.tif"))

# Import tile h08v05 from 2021-02-16
nl_h08v05_16 <- read_stars(here("data", "VNP46A1","VNP46A1.A2021047.h08v05.001.2021048091106.tif"))

# Import tile h08v06 from 2021-02-16
nl_h08v06_16 <- read_stars(here("data", "VNP46A1","VNP46A1.A2021047.h08v06.001.2021048091105.tif"))

```

### Roads data import
This vector data is from OSM, processed with the help of a third party site (CITE) to produce a shapefile of all Texas highways. This data contains 6,085 observations with 11 variables. 
```{r}
# Import roads data with SQL query to filter for only highways (use st_read)
roads <- read_sf(here("data", "gis_osm_roads_free_1.gpkg"), query = "SELECT * FROM gis_osm_roads_free_1 WHERE fclass='motorway'")

```

### House data import
This vector data is also from OSM, processed with the help of a third party (CITE) to produce a shapefile of houses in the Houston metropolitan area. This data contains 475941 observations with 6 variables. 
```{r}
# Import house data with SQL query to filter for selected building types (use st_read)
houses <- read_sf(here("data", "gis_osm_buildings_a_free_1.gpkg"), query = "SELECT * FROM gis_osm_buildings_a_free_1 WHERE (type IS NULL AND name IS NULL) OR type in ('residential', 'apartments', 'house', 'static_caravan', 'detached')")

```

## Socieconomic data import
This data is from  U.S. Census Bureau’s American Community Survey (CITE) 2019 census. This Geodatabase (gdb) contains data for Texas. I will be importing two layers from the gdb. 
```{r}
# Import socioeconomic geometry layer
socioeco_geom <- read_sf(here("data", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"), layer = "ACS_2019_5YR_TRACT_48_TEXAS")

# Import socioeconomic income layer 
socioeco_income <- read_sf(here("data", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"), layer = "X19_INCOME")

# View layers with 'st_layers' (commented out for now)
#st_layers(here("data", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"))

```

## Data Exploration

### CRS Matching 
Before diving into the data, I want to see what CRSs the data is currently at. 

```{r}
# ......Check CRS of individual datasets......

# Create list of the datasets that have geometries
datasets <- list(houses, roads, socioeco_geom, nl_h08v05_07, nl_h08v06_07, nl_h08v05_16, nl_h08v06_16)

# Loop through each dataset and print its CRS

for (i in seq_along(datasets)) {
  epsg <- st_crs(datasets[[i]])$epsg
  print(epsg)
}

```

### Check for Invalid geometries
Check for invalid geometries in our three datasets (roads, houses, socieoeco_geom) that contain a geometry type (e.g, multiploygon). 
```{r}
#......Checking for invalid geometries......

# Create a dataset with only the data that contains geometries
datasets_geom = list(roads, houses, socioeco_geom)

# Check if there are invalid geometries 
for (i in seq_along(datasets_geom)) {
  which(!st_is_valid(datasets_geom[[i]]))
  print("There are NO invalid geometries")
}
```

## Night Lights

### Merge tiles 
In this section, I merged the tiles for February 7th, 2021 and February 16, 2021. 
```{r}
#......Night light change ......

# Merge night light for 2021-02-07
nl_07 <- st_mosaic(nl_h08v05_07, nl_h08v06_07)

# Merge night light for 2021-02-16
nl_16 <- st_mosaic(nl_h08v05_16, nl_h08v06_16)

# Confirm class is
print(class(nl_07))
print(class(nl_16))

# Clear up environment for efficiency 
rm(list = c("nl_h08v05_07", "nl_h08v06_07"))
rm(list = c("nl_h08v05_16", "nl_h08v06_16"))

```

## Map: Night Lights 
In this section, I mapped the night lights for February 7th and 16th. In order to do so, I created a bounding box for the Houston area to crop the night light data with.  
```{r}
# Create bounding box for Houston area 
bbox_houston <- st_bbox(c(xmin = -96.5, xmax = -94.5, ymax = 30.5, ymin = 29),crs = st_crs(nl_07)) # Match CRS to night light data

# Crop Feb 7th night light data
nl_07_crop<- st_crop(nl_07, bbox_houston)

# Crop Feb 16th night light data
nl_16_crop<- st_crop(nl_16, bbox_houston)

# Verify that 'nl_07' and 'nl_16' were cropped to Houston area

if(ncell(nl_07) == nrow(nl_07_crop)) {
  warning("data did not crop!")
} else {
  print("data cropped!")
}

if(ncell(nl_16) == nrow(nl_16_crop)) {
  warning("data did not crop!")
} else {
  print("data cropped!")
}

# Clear up environment for efficiency 
rm(list = c("nl_07"))
rm(list = c("nl_16"))

```

```{r}
#......Mapping 2021-02-07......

nl_07_map <- tm_shape(nl_07_crop) + # Map cropped data
  tm_raster(palette = "Purples", 
            breaks = c(0, 100, 200, 300, 400, 500, 600, 700, 800, 900, 1000), # custom breaks
                       title = "Light Intensity (nW/cm^2/sr)") +
  tm_title("Houston, Texas Night Lights 02-07-2021", fontface = "bold") +
  tm_title("Before storm (02-16-2021)", size = 0.8) +
  tm_layout(bg.color = "white", legend.show = TRUE) +
   tm_compass(size = 0.8) +
  tm_scalebar(size = 0.5)

# View map
nl_07_map

```
```{r}
#......Mapping 2021-02-16......

nl_16_map <- tm_shape(nl_16_crop) + # map cropped data
  tm_raster(palette = "Purples",
            breaks = c(0, 100, 200, 300, 400, 500, 600, 700, 800, 900, 1000), # custom breaks 
            title = "Light Intensity (nW/cm^2/sr)") + 
  tm_title("Houston, Texas Night Lights", fontface = "bold") +
  tm_title("After storm: 02-16-2021 ", size = 0.8) +
  tm_layout(bg.color = "white", legend.show = TRUE) +
  tm_compass(size = 0.8) +
  tm_scalebar(size = 0.5) 

# View map
nl_16_map

```

### Find blackouts 
Find blackout regions by defining the blackout regions where more than 200 nW cm-2sr-1 experienced a blackout and those less than 200 nW cm-2sr-1 did not experience a blackout. 
```{r}
#......Vector difference mask ......

# Create conditional where TRUEs are regions that experienced a blackout 
blackout <- (nl_16 - nl_07) > 200

# Reassign FALSEs as NA
blackout[blackout == FALSE] <- NA

# Turn into vector object 
blackout <- st_as_sf(blackout)

# View your new vector dataframe 
#head(blackout)

# Check for invalid geometries
if (any(!st_is_valid(blackout))) {
  blackout <- st_make_valid(blackout)
} else if (nrow(blackout) == 0) {
  print("No invalid geometries found.")
} else {
  print("All geometries valid!")
}


```

```{r}
#......Crop 'Blackout' mask to Houston area ......

# Crop your sf object to Houston bbox made previously
blackout <- st_crop(blackout, bbox_houston)

# Re-project to EPSG:3083 (NAD83 / Texas Centric Albers Equal Area)
blackout <- st_transform(blackout, crs = "EPSG:3083")

# Test that CRS re-projected
test_that("Test 1", {
  expect_equal(st_crs(blackout)$epsg, 3083)
}) 
```

## Map: Changing in night light intensity
```{r}
#......Mapping Blackout regions......

blackout_map <- tm_shape(blackout) +
  tm_polygons(col = "#2f0873") +
  tm_title("Blackout Regions February 2021 Texas Power Crisis", fontface = "bold")+
  tm_title("source: NASA VIIRS", size = 0.5) +
  tm_layout(bg.color = "white") +
  tm_basemap("CartoDB.PositronNoLabels") +
  tm_compass(size = 0.8) +
  tm_scalebar(size = 0.5)

# View map
blackout_map


```

## Highways
In this section, I will exclude highways from the blackout mask because they may have experienced night light changes not associated with the severe storms. 

```{r}
#......Create buffer of 200 meters ......

# The CRS of 'roads' needs to be transformed to the Texas CRS
roads <- st_transform(roads, crs = "EPSG:3083")

# Check the units of 'roads'
units_roads <- st_crs(roads)$units

if (units_roads != "m") {
  warning("Units are NOT in meters!")
} else {
  message("Units are in meters!")
}

# Identify areas within 200 m of all highways
highway <- st_buffer(roads, dist = 200) %>% 
  st_union()

# Check for invalid geometries 
which(!st_is_valid(highway)

# Check that buffer is a polygon 
st_geometry_type(highway)


```

```{r}
#......Areas further than 200 m......

blackout_highways <- st_difference(blackout, highway) 

```


## Buildings
In this section I found the buildings in Houston, TX affected by the blackout. I used the function 'st_join' by intersection to find the houses all within the 'blackout_highways' multipolygon data frame that experienced a blackout.
```{r}
#......Find buildings affected by blackout ......

# The CRS of 'roads' needs to be transformed to the Texas CRS
houses <- st_transform(houses, crs = "EPSG:3083")

# Join blackout highways and houses 
buildings_blackout <- st_join(houses, blackout_highways, join = st_intersects, left = TRUE)

# View the class of the new data frame 
print(class(buildings_blackout))

## FIND estimate of houses that experienced blackout

```

### Map: Homes in Houston that lost power 
```{r}
tm_shape(buildings_blackout) +
  tm_polygons(col = "#2f0873") +
  tm_title("Blackout Regions February 2021 Texas Power Crisis", fontface = "bold")+
  tm_title("source: NASA VIIRS", size = 0.5) +
  tm_layout(bg.color = "white") +
  tm_basemap("CartoDB.PositronNoLabels") +
  tm_compass(size = 0.8) +
  tm_scalebar(text.size = 0.5)

```


## Census Tracts 
In this section I investigated which census tracts experienced blackouts or no blackouts. 

Some important information for this analysis: 
- socioeco_geom = census tracts geometries
- socioeco_income = median income  

Some data wrangling before...
```{r}

# rename GEOID column in socieco_geom 
socioeco_geom <- socioeco_geom %>% 
  select(-GEOID) %>% 
  rename(GEOID = GEOID_Data)

# Join layers of socioeco and select columns relevant for analysis
socieoco <- left_join(socioeco_geom, socioeco_income, by = "GEOID") %>% 
 select(GEOID, B19013m1, Shape) %>% 
  rename(geom_soc = Shape)

# Check CRS of 'socieoco' and 'buildings_blackout'
if(st_crs(socieoco) == st_crs(buildings_blackout)) {
  print("coordinate reference systems match")
} else{
  print("coordinate reference systems do NOT match")
} 

# Transform CRS 
if(st_crs(socieoco)$epsg != 3083) {
  print("CRS not for Texas! Transforming...")
  socieoco <- st_transform(socieoco, crs = "EPSG:3083")
}

# Update geom name and select for geometry on in buildings_blackout
buildings_blackout <- buildings_blackout[, "geom"]


```


Finding what census tracts that did not experience a blackout.
```{r}
#.....Census tracts that experienced blackout ......

soc_blackout <- st_join(socieoco,buildings_blackout, left = TRUE) 

# make column with median incomes called (income_blackout)

soc_no_blackout <- st_disjoint(socieoco[socieoco,buildings_blackout], left = TRUE) 

# make column with median incomes called (income_no_blackout)

# rejoin data set 

# create new variable that tells where they experienced blackout or not, group by this variable 

soc_aft_houses %>% 
  mutate()


# data wrangling to summarize median income by census tract 
# In META data - B19051e1
soc_aft_houses %>% 
  group_by(B19013m1) %>% 
  aggregate(B19013m1))


# the NAs are those that did not experience blackout 

# to find those that are not expe blackouts, go back to joining them 

# make a column called 'experienced blackout' vs 'did not experience blackout (NAs)'

# make box plot of median income

# plot map just cause 
```


